---
title: "team.Rmd"
author: 'Chase Weida; Github: cweida; Bernard Fay: Github: bj99fay'
date: "3/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tidyverse)
```

## Cleaning
```{r}
#defense
defense <- read_excel('cyclonesFootball2019.xlsx', sheet='Defensive')
defense <- data.frame(defense)
defense$Name <- as.factor(defense$Name)
defense$Opponent_Opponent <- as.factor(defense$Opponent_Opponent)

defense <- defense %>% mutate_at(.vars=c(3:11), .funs= list(as.numeric))
defClean <- defense
str(defClean)

#offense
offense <- read_excel('cyclonesFootball2019.xlsx', sheet='Offensive')
offense$Name <- as.factor(offense$Name)
offense$Opponent_Opponent <- as.factor(offense$Opponent_Opponent)
offense <-separate(offense, "Passing_CMP-ATT", into=c("Passing_CMP", "Passing_ATT"), sep="-\r\n ")
offense <- mutate_at(offense, .vars=c(3:13), .funs=list(as.numeric))

offClean <- offense
str(offClean)

#biography
biography <- read_excel('cyclonesFootball2019.xlsx', sheet='Biography')
biography <- data.frame(biography)

biography <- biography %>% separate(Height, sep = "-", into = c('Ft', 'inch'), remove = TRUE)  
biography$Ft <- as.numeric(biography$Ft)
biography$inch <- as.numeric(biography$inch)

biography$Height <- biography %>% transmute(inches = (Ft * 12) + inch)
biography <- biography %>% select(-c(3,4))
biography <- biography %>% mutate_at(.vars = c(1,2,4,5,6), .funs = list(as.factor)) %>% mutate_at(.vars = c(3), .funs = list(as.numeric))
bioClean <- biography
str(bioClean)
```

## Tidying(1)
### 1.
```{r}
dat1 <- defClean %>% pivot_longer(cols=c(3:11), names_to="Statistic")
```

### 2.
```{r}
dat1 %>% ggplot(aes(x=as.numeric(value))) + geom_bar(stat="count") + facet_wrap(~Statistic) + xlab("value")
```

There are several defensive skills that are rare. Pass_PB, Pass_QBH, Tackles_Sack, Tackles_TFL, Turnover_FF, Turnover_FR, and Turnover_INT are the rare defensive skills because very few players have non-zero values in these areas.

### 3.
```{r}
dat1 %>% head()
dat2 <- dat1 %>%
        filter(Opponent_Opponent %in% c("Kansas State", "West Virginia") & Statistic == "Tackles_Solo") %>%
        select(Name, Opponent_Opponent, value) %>%
        pivot_wider(names_from = Opponent_Opponent, values_from = value)
dat2$`West Virginia`[is.na(dat2$`West Virginia`)] <- 0
dat2$`Kansas State`[is.na(dat2$`Kansas State`)] <- 0
dat2 %>% ggplot(aes(x=`West Virginia`, y=`Kansas State`)) + geom_point() + 
        geom_text(aes(label=ifelse(`West Virginia`>2, as.character(Name),'')), hjust=0, vjust=0) +
        geom_text(aes(label=ifelse(`Kansas State`>2, as.character(Name),'')), hjust=0, vjust=0)
```

Iowa State had better defense against Kansas State as there were several players with more than 2 tackles in that game, compared to only five players with >2 tackles in the West Virginia game.

## Tidying (2)
### 1
```{r}
biography <- bioClean %>% separate(Hometown, se = ", ", into = c("Home City", "Home State"))
biography[c(1:3),c("Home City", "Home State")]
```

### 2
```{r}
biography %>% add_count(`Home State`, name = "Players in State") %>% select(`Home State`, `Players in State`) %>% distinct()
```
Here one can see that the players in the data set come from 19 total states. 35 players come from Iowa while the next greatest number of players come from Florida. 

## Joining
### 1
```{r}
offJoin <- biography %>% inner_join(offClean, by = 'Name')
offJoin[,c('Name', "Home City", "Home State")] %>% distinct()
```


### 2
```{r}
ggplot(data = offJoin, mapping = aes(Weight, Receiving_YDS)) + geom_point() +geom_smooth(method = 'lm', se = FALSE)
```

As seen in the plot above, there does not exist a very strong relationship between the weight of the player and the receiving yards of the player. As the weight of the player increases the expected number of receiving yards neither increases nor decreases. 

### 3
```{r}
offDefJoin <- offClean %>% full_join(defClean, by = c("Name", "Opponent_Opponent"))

purdy <- offDefJoin %>% filter(Name == "Purdy, Brock") %>% select(Opponent_Opponent, Passing_YDS, Rushing_YDS)

columns <- c("Opponent_Opponent", "Receiving_YDS", "Rushing_YDS",
             "Tackles_Sack", "Turnover_FR", "Turnover_INT")

offDefJoin <- offDefJoin[,columns]
offDefJoin <- offDefJoin %>% group_by(Opponent_Opponent) %>%
                summarise(Receiving_YDS = sum(Receiving_YDS, na.rm = TRUE),
                          Rushing_YDS   = sum(Rushing_YDS, na.rm = TRUE),
                          Tackles_Sack  = sum(Tackles_Sack, na.rm = TRUE),
                          Turnover_FR   = sum(Turnover_FR, na.rm = TRUE),
                          Turnover_INT  = sum(Turnover_INT, na.rm = TRUE))
offDefJoin <- offDefJoin %>% mutate(YDS = Receiving_YDS + Rushing_YDS) %>% 
                             mutate(DEF = Tackles_Sack + Turnover_FR + Turnover_INT)

offDefJoin$YDS <- offDefJoin$YDS/mean(offDefJoin$YDS)
offDefJoin$DEF <- offDefJoin$DEF/mean(offDefJoin$DEF)
offDefJoin <- offDefJoin %>% mutate(YDSplusDEF = (YDS + DEF)/2)

purdy$Passing_YDS <- purdy$Passing_YDS/mean(purdy$Passing_YDS)
purdy$Rushing_YDS <- purdy$Rushing_YDS/mean(purdy$Rushing_YDS)
purdy <- purdy %>% mutate(YDS_Total = (Passing_YDS + Rushing_YDS)/2)

purdy %>% ggplot(aes(x=Passing_YDS, y=offDefJoin$YDSplusDEF)) + geom_point() +geom_smooth(method = 'lm', se = FALSE) + xlab("Purdy Total YDS") + ylab("Team YDS + DEF")
```

After standardizing the values (where 1 is average), there does not appear to be a correlation between Brock Purdy's performance and the team's overall perforamce. Purdy's performance was based on how many yards he passed and rushed for in a game. The team's performance was based on how many receiving and rushing yards, sacks, and turnovers forced the team had in a game.


### 4
What state has the highest net turnovers (forced/recovered - given up)?
```{r}
offDefJoin2 <- offClean %>% full_join(defClean, by = c("Name", "Opponent_Opponent"))
allJoin <- biography %>% inner_join(offDefJoin2, by = "Name")
allJoin <- allJoin[,c("Home State", "Passing_INT", "Turnover_FF", "Turnover_FR", "Turnover_INT")]

allJoin[is.na(allJoin)] <- 0
allJoin <- allJoin %>% group_by(`Home State`) %>% summarize_each(list(sum))
allJoin <- allJoin %>% mutate(`Net Turnover` = Turnover_FF + Turnover_FR + Turnover_INT - Passing_INT)
allJoin %>% ggplot(aes(x=`Home State`, y=`Net Turnover`)) + geom_bar(stat="identity") + coord_flip()
```

Iowa had the highest net turnovers. However, the data set did not contain information on fumbles when rushing or receving. If this data was included, the results would likely be different because the only source of turnovers on offense are passing interceptions, which are all made by the Quaterback Brock Purdy.
